Introduction_outline

### 2019-01-25 Write_up/Writing/Introduction_outline.Rmd
updated

Climate change as a result of anthropogenic activities is now established as a global concern for both scientific and political importance. As the human population grows, the rate of global warming increases, resulting in profound effects on marine and other ecosystems. Climbing temperatures create a host of additional changes to marine systems, such as rising sea levels, increased ocean stratification, decreased sea-ice extent, and altered patterns of ocean circulation and precipitation [@doney]. Furthermore, direct effects on ocean temperature changes and chemistry may alter the functioning and behaviour or marine organisms. In turn, these changes lead to a change in species integration and trophic pathways.

Significant research studies accept the climate change and variability have great impacts on marine ecosystems [@stenseth; @harley; @hoegh]. The Intergovernmental Panel on Climate Change (IPCC) predicts future projections with a mean increase in air temperature of 1.1°C by the 2020’s and 1.8°C by the 2040’s as a result of greenhouse gas concentrations [@solomon]. A large amount of this heat will be absorbed by the world’s ocean. In current years, as a result of changes in air temperature, the heat content of the ocean has increased largely. Given ocean warming in association with their physical changes such as sea level rise and declines in sea ice, marine life is expected to intensify in future decades. Global warming represents distinct warming and cooling periods throughout the 20th and 21st century (IPCC, 2007). Global warming however is not evenly distributed across the global ocean thus presenting regions warming much faster or slower than then the global average [@levitus; @santos]. In particular, the Atlantic Ocean contributes most to the increase of the heat content [@nerem; @levitus].

As a result of climate change, research studies provide significant evidence on the increase in warming of the ocean surface. This warming slowly propagates down through the water column and in turn increases the thermal stratification [@gruber]. Strong stratification limits the depth at which water is upwelled and hence controls the amount of nutrients brought up [@chhak; @jacox2011; @jocox2016]. The interplay between upwelling winds and stratification determine the amount of nutrients present within the system and thus is strongly linked to biological productivity. Globally, as a result of climate change, stratification of the water-column between 0 and 200 m has increased by an estimated 4% between 1971 to 2010 [@Stocker]. Increased ocean stratification as a result of surface warming may reduce the ventilation of deep-water masses. As the deep water is being up welled to the surface, decreases in oxygen and pH levels are predicted along with an increase in nutrient levels [@Rykaczewski).  Generally, regional ocean climate oscillations like the Benguela Niño for the Benguela Current System impose considerable variability on temperatures and complicate attempts to assess long-term trends [@demarcq]. It is predicted that these climate oscillations may increase the variance with climate change [@timmermann; @kuzmina; @sydeman] which will further result in unusual trends.

@@

@@

With global warming continuing at a rapid pace the amount of greenhouse gases within the atmosphere are going to increase in the future. This increase will undoubtedly have an impact on the world’s oceans and the marine life within, both directly and indirectly. The oceanic process of coastal upwelling is one aspect that is likely to be affected as changes in global temperature will affect temperature and pressure gradients, and in turn affect alongshore winds which are responsible for driving this process within EBUS regions. Because upwelling is important for productivity within coastal systems, a decrease in this process will negatively impact on marine fauna and flora, and on human activities such as fishing. To assess changes in upwelling events, here I aim to use SST, wind, chlorophyll data etc. to monitor durations and intensities of upwelling within one of the EBUS.....

@@

@@

Extreme thermal events have a range of broad scale definitions, but they can each be classified into two categories, namely, cold-spells [e.g. @gunter; @lirman; @boucek] or heatwaves [e.g. @gordon; @stott; @perkins]. These events occur rapidly, and unexpectedly and they profoundly impact on species diversity, distributions, growth and survival where they take place. These events are often undetected but studies have found that extreme events have occurred as recently as approximately four decades ago. Jentsch et al. (2007) were the first to discuss this mode of climate change and brought it to focus. Steps have since been taken towards understanding the drivers, impacts and consequences of these events. Subsequently, research has predicted an increase in the occurrence of these extreme events as a result of long-term warming of the planet via the process of global warming (REFERENCE). Extreme events are often associated with toxic algal blooms, shifts in species distribution, and the bleaching of coral reefs [@mccabe; @harley; @schoepf].

Little is known about extreme events occurring within the South African region. Very little South African researchers spend their time and careers on studyig extreme events within this region. During the 1980s, a study done by Birkett and Cook was the only study done where they anaylsed the impact of extreme warming events on the life cycle of white muscles,Donax serra Röding. However, more recently studies done by Schlegel et al, (2017,2018) examined the relationship between the nearshore and offshore occurances of marine heatwaves and cold-spells as well as the influence of predominant atmosphetic and oceanic patterns during Coastal Marine Heatwaves. It is now becoming accepted that extreme events are a threat to several ecosystem [@thompson].

### Cold spells

- Definition
- Impact
-Major cold spells

## Aims and approach:
In this thesis I aim to…

>>>>

Climate change as a result of anthropogenic activities is  an established global concern of both scientific and political importance. As the human population grows, the rate of global warming increases, resulting in profound effects on marine and other ecosystems. Climbing temperatures create a host of additional changes to marine systems, such as rising sea levels, increased ocean stratification, decreased sea-ice extent, and altered patterns of ocean circulation and precipitation [@doney]. Furthermore, direct effects on ocean temperature changes and chemistry may alter the functioning and behaviour or marine organisms. In turn, these changes lead to a change in species integration and trophic pathways.

Significant research studies accept the climate change and variability have great impacts on marine ecosystems [@stenseth; @harley; @hoegh]. The Intergovernmental Panel on Climate Change (IPCC) predicts future projections with a mean increase in air temperature of 1.1°C by the 2020’s and 1.8°C by the 2040’s as a result of greenhouse gas concentrations [@solomon]. A large amount of this heat will be absorbed by the world’s ocean. In current years, as a result of changes in air temperature, the heat content of the ocean has increased largely. Given ocean warming in association with their physical changes such as sea level rise and declines in sea ice, marine life is expected to intensify in future decades. Global warming represents distinct warming and cooling periods throughout the 20th and 21st century (IPCC, 2007). Global warming however is not evenly distributed across the global ocean, and thus several regions are warming faster or slower than then García-Reyes et al., 2015 the global average [@levitus; @santos]. In particular, the Atlantic Ocean contributes most to the increase of the heat content [@nerem; @levitus].

As a result of climate change, research studies provide significant evidence relating to the increase in the warming of the ocean surface. This warming slowly propagates down through the water column and in turn increases the thermal stratification [@gruber]. Strong stratification limits the depth at which water is upwelled and hence controls the amount of nutrients brought up [@chhak; @jacox2011; @jocox2016]. The interplay between upwelling winds and stratification determines the amount of nutrients present within the system and is therfore strongly linked to biological productivity. Globally, as a result of climate change, stratification of the water-column between 0 and 200 m has increased by an estimated 4% between 1971 to 2010 [@Stocker]. Increased ocean stratification as a result of surface warming may reduce the ventilation of deep-water masses. As the deep water is being up welled to the surface, decreases in oxygen and pH levels are predicted along with an increase in nutrient levels [@Rykaczewski).  Generally, regional ocean climate oscillations like the Benguela Niño for the Benguela Current System impose considerable variability on temperatures and complicate attempts to assess long-term trends [@demarcq]. It is predicted that these climate oscillations may increase the variance with climate change [@timmermann; @kuzmina; @sydeman] which will further result in unusual trends.

@@

Climate change has severe ecological consequences on both plants and animals (Bradley et al. 1999; Walther et al. 2002). For example, in the past two decades, sea surface temperature has risen in many regions. Seasonal rise in temperature can act as an important cue in species survival, reproduction and abundance (Cushing 1982; Cushing 1990). The differing responses of species as a result of climate change are important to understand as these changes affect, among many things, the trophic dynamics and fisheries yield of several regional ecosystems (Murawski 1993).

Understanding the relationship between global warming and climate change as a result of anthropogenic activity is a vast and challenging process. In a bid to further this understanding, Rockström et al. (2009) introduced the concept of planetary boundaries. According to this concept, within each of these boundaries exists a level of safety that should not be exceeded. For example, the planetary boundary for biodiversity loss, has been surpassed (Steffen et al., 2015). Similarly, the safety boundary for warming of the globe by greenhouse gas emission has also been surpassed. The idea of planetary boundaries is essential in that it helps our understanding of the threats we cause via the continued destruction of the natural environment. Consequently, studies of anthropogenic influences on the ecosystem are generally divided into land or sea. Terrestrial ecosystems are more familiar to the general populace; whereas marine ecosystems are associated with high levels of primary production (Chavez et al., 2011) and are highly at risk as a result of climate change (Österblom et al., 2017).  This has been proven to decrease primary production thus affecting the trophic levels and leading to trophic turbulence and destruction of marine fauna (Lewandowska et al., 2014; Dell et al., 2014, McCauley et al., 2015).

Climate change similarly affects coastal species such as oysters, corals and, sea grass. Lord (2017) also found evidence that suggests that a change in climate as a result of rising sea temperatures may increase the spread of invasive species and increase the chances of an outbreak of diseases (Altizer et al., 2013). Research by Mead et al (2013) found that coastal ecosystems are at high risk as a result of climate change, later, Whitfield et al (2016) confirmed that coastal bio-diversity have been observed to be affected. Given the importance of primary productivity, it was determined that the Benguela upwelling systerm within the South African region should become the focus of this thesis. Here the increase in global temperatures are one of the main focus points along with the chlorophyll concentration. By making use of these two measurements, I am able to analyses how chlorophyll concentration vary with a change in temperature.

@@

With global warming continuing at a rapid pace, the amount of greenhouse gases within the atmosphere are going to increase in the future. This increase will undoubtedly have an impact on the world’s oceans and the marine life within, both directly and indirectly. The oceanic process of coastal upwelling is one aspect that is likely to be affected as changes in global temperature will affect temperature and pressure gradients, and in turn affect alongshore winds which are responsible for driving this process within EBUS regions. Because upwelling is important for productivity within coastal systems, a decrease in this process will negatively impact on marine fauna and flora, and on human activities such as fishing. To assess changes in upwelling events, here I aim to use SST, wind, and chlorophyll data to monitor the durations and intensities of upwelling events within the Benguela EBUS and identify wether this resulted in an extreme anomalous cold event. Atmospheric-ocean coupling is known to be affected by climate change, likely changing wind patterns and leading to upwelling favourable conditions (García-Reyes et al., 2015). Here we test if an increase in upwelling may lead to a detection of more MCSs simply because these cold temperatures caused by aseasonal upwelling will reflect in historic climatologies.

@@

Benguela Niños are defined as anomalous warm events occurring between the southward flowing Angola current and the Benguela upwelling system off south western Africa (Shannon et al. 1986). These warm events profoundly impact on fisheries and the climate of this region. They are proven to induce unexpected rainfall events and drastically influence fish abundance and distribution (Rouault et al. 2003, Boyer et al. 2001). Evidently, Benguela Niños are expressed as regions of abnormal, persistent high sea surface temperatures. Many researchers often associate the equatorial interannual variability pattern in the Atlantic to the ENSO phenomenon found within the Pacific region (Sutton et al. 2000; Zebiak 1993). Benguela Niños have been recorded to occur in 1934, 1949, 1963, 1984 and 1995 (Shannon et al. 1986; Gammelsrød et al. 1998). Picaut (1985) suggested that eastern tropic oceans are governed by remote wind stress effects through equatorial wave dynamics. Anomalies in trade winds may result in Kelvin waves that propagate eastward along the equator, inducing a deepening or a lifting of the thermocline. It is possible that equatorial Kelvin waves may play a role in the evolution of Benguela Niños.

@@

Extreme thermal events have a range of broad scale definitions, but they can each be classified into two categories, namely, cold-spells [e.g. @gunter; @lirman; @boucek] or heatwaves [e.g. @gordon; @stott; @perkins]. These events occur rapidly, and unexpectedly and they profoundly impact on species diversity, distributions, growth and survival where they take place. These events are often undetected, but studies have found that extreme events have occurred as recently as approximately four decades ago. Jentsch et al. (2007) were the first to discuss this mode of climate change and brought it to greater focus. Steps have since been taken towards understanding the drivers, impacts and consequences of these events. Subsequently, research has predicted an increase in the occurrence of these extreme events as a result of long-term warming of the planet via the process of global warming [@mccabe; @harley; @schoepf].

Little is known about extreme events occurring within the South African region. Few South African researchers have specifically focussed on investigating this topic, and published literature are sparse. During the 1980’s however, Birkett and Cook analysed the impacts of extreme warming events on the life cycle of white muscles, Donax serra Röding. Futher studies on extreme events occurred within the Florida Reef Tract, where anomolously cold events resulted in the mass mortality of coral reefs and fish. Acropora cervicornis, was found to be particularly susceptible to low temperatures and experienced extensive mortality (Shinn 1966; Davis 1982). While these phenomenons are not as common as warming events, they have a dramtic and long-lasting effect on coral communities. However, studies in January 2010 reveal that the Florida Reef Tract experienced a cold-water anomaly, marine cold spells (MCSs) with temperatures below 16 ◦ C recorded for up to 14 days (Lirman et al. 2011).  Species population distributon are often set by cold distribution boundaries.  More recently, Schlegel et al, (2017, 2018) examined the relationship between the nearshore and offshore occurrences of marine heatwaves and cold-spells as well as the influence of predominant atmospheric and oceanic patterns during Coastal Marine Heatwaves. It is now becoming accepted that extreme events are a threat to several ecosystems [@thompson]. In a warming planet one would assume that the occurrence and intensity of the severity of marine cold spells (MCSs) are likely to decrease. This thesis aims to address this assumption.

## Aims and approach:

The warming of the global climate as a result of anthropogenic activity, has negatively influenced both terrestrial and marine ecosystems, and ultimately inhibits humanity and ecological functioning (McCauley et al., 2015). This climate variability, manifested as increases in frequency and duration of extreme events is regarded as a great threat (Thompsonet al., 2013). Very few studies have addressed the occurrence of cold temperature spells or cold events within the Benguela region. This investigation examines recent trends and variation in the frequency and duration of cold spells occurring in this region. Due to the lack of in situ collected data, cold spells are mainly examined between the period 2014-2017 for four sites along the South African coastline falling within the Benguela upwelling cell. Here we aim to assess whether the intensity and duration of these extreme events may show any changes in frequency and intensity on a seasonal basis. In this research project I make use of various satellite derived observations of sea surface temperature (SST) and wind stress to further investigate the Benguela upwelling system along the South African coastline, whilst considering cold events and warm events occurring within this region
